\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-11-05 paper.tex
processing of review comments

% additional packages and utility commands

\usepackage{float}

\newcommand{\TODO}{\textbf{\color{red}TODO}}

% \usepackage{flushend}

% while preparing, linenumbers come in handy
\usepackage{lineno}
\setlength\linenumbersep{1mm}
\linenumbers

% until we settle on the final name ;-)
\usepackage{xspace}
\newcommand{\NAME}{id-foo\xspace}

% http://tex.stackexchange.com/questions/299/how-to-get-long-texttt-sections-to-break
\newcommand*\justify{%
  \fontdimen2\font=0.4em% interword space
  \fontdimen3\font=0.2em% interword stretch
  \fontdimen4\font=0.1em% interword shrink
  \fontdimen7\font=0.1em% extra space
  \hyphenchar\font=`\-% allowing hyphenation
}

% http://en.wikibooks.org/wiki/LaTeX/Customizing_LaTeX
\newcommand{\ttt}[1]{\texttt{\justify{#1}}}

\usepackage{algpseudocode}
% custom Let command
\newcommand*\Let[2]{\State #1 $\gets$ #2}
\renewcommand{\algorithmicforall}{\textbf{for each}}
\let\ForEach\ForAll

\usepackage{listings}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{graphs}
\usetikzlibrary{trees}
\usetikzlibrary{arrows}

\usepackage{tikz-3dplot}

\include{tikz}
\include{foo-lang}

\usepackage[autostyle]{csquotes}

@@

% to avoid syntax error highlighting - e.g. foo's not js ;-)
\expandafter\def\csname PY@tok@err\endcsname{}

@@

\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Christophe
Huygens, Danny Hughes, Wouter Joosen}%
\IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
3001 Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be}}

@@

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns, anomalies,\dots that can indicate malicious
behavior. But supporting these intrusion detection algorithms imposes
significant overhead on resource-restricted devices such as those targeting the
Internet of Things. Fusion of the algorithms, to optimize the use of the shared
resources, can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce \NAME, a framework for the
development of efficient intrusion detection systems. It consists of a domain
specific language, to formally describing intrusion detection algorithms on a
functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

The internet of things (IoT) presents a great potential to positively influence
our daily work and life through domotics, assisted living, e-health and
enhanced learning\cite{atzori2010internet}. But with this potential the IoT
also presents a significant threat: by opening our smart homes and personal
data to all these interconnected devices, we also open them to everyone who is
able to break the virtual locks that protect them. IoT devices typically have
limited batteries, processing power and memory. These properties make it hard
to add any additional service, including security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally, after detecting a malicious
patter, alerting the firewall\cite{denning1987intrusion}.

% manet: local ids -> localize -> limited resources

In wireless networks of resource constraint devices, which make up an
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic\cite{mishra2004intrusion}. If the
outer security perimeter diminishes, every device has to implement its own
lines of defense. But making the IDS a local service on each device requires
local resources, which are the scarce commodity to IoT devices.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms\cite{valero2012di} and offer the required basic
functional components to implement them\cite{krontiris2008lidea}. The first
aspect that is missing, is that they don't offer a way to optimize the usage of
resources. A solution that supports the integration of multiple algorithms on
IoT devices needs to consider the impact on their resources and avoid
accumulating their impact.

Secondly, they don't support systematic reuse to create different
configurations for different devices, both dealing with hard- and software.
Addressing this in a transparent and automated way is important for any kind of
development in this area.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
for patterns in the implementation of these classes of algorithms, details some
of the ideas behind FCF and identifies the different aspects of the problem
space. Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

@@

Research into ID in wireless networks typically focuses on two major topics
that complement the architecture: the devices as single entities and the
network as a group of such devices. Both are important because the group cannot
make a decision without members that detect malicious behavior and a
group-based decision is often needed because nodes can easily miss out on
certain events that could indicate intrusions, due to their wireless and
not-always-on nature.

These two topics lead to the first two classes: detecting intrusions and
cooperative decision making. We also identify software attestation as a third
class. Although it is a combination of both other classes, its focus is not on
the detection of intrusions, but on the effect of an intrusion on the system
itself.

\subsection*{Detecting Intrusions}
\label{detection}

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors agree with this topology and add the
notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories seem promising, the authors admit that the impact on
resource-constrained devices will be too much. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Anomaly detection] considers a baseline profile and evaluates all
  activity according to this profile. It tries to flag aberrant behavior as a
  possible intrusion, but this is no exact science. The baseline profile can
  evolve over time, especially in wireless and mobile environments. Single
  events should therefore not result in absolute decisions.

  \item[Signature or misuse detection] starts with known patterns of attacks
  and tries to find evidence in the actions of and communication with a device.
  Because attacks often use valid ways to communicate with services on the
  devices, this can also produce false positives and offers no black and white
  certainty.

@@

\subsection*{Cooperative Decision Making}
\label{coorperative}

These taxonomies focus on the act of detecting intrusions themselves. In a
highly distributed environment, an orthogonal categorization, dealing with the
distributed nature of the network and its ways of communicating, may be even
more important.

A group is often stronger than the sum of its individual components. Because
not all devices are always actively participating in the network, they can miss
clues that would otherwise lead them to detect intrusions. It is hardly
impossible for a single device to detect a complex attack by itself. Therefore
a second important research topic consists of cooperative algorithms to combine
information from single identities into a group-based decision about alleged
intrusions. We identify two ways to construct cooperative decision making and
focus on the way participants exchange information: broadcasting and
interactive.

@@

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol\cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, the
  participants can make a distributed decision about the jointly identified
  intruder.

@@

\subsection*{Software Attestation}
\label{subsection:attestation}

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. It offers algorithms for exchanging
information about the software that is running on a device and allows to
identify devices with modified content, that no longer can pass attestation. A
good example is SAKE\cite{seshadri2008sake}.

Software attestation is hard and details can cause an algorithm to fail. The
general feeling is that even in the best conditions, there is always a way to
circumvent the most ingenious algorithm to perform software-based attestation.

Within the context of this paper, we will not consider software attestation as
a special class and exclude it from our scope, only focusing on detecting
intrusions and distributed decision making. Figure \ref{fig:classification}
visually summarizes this dual classification.

@@

Given this classification, we now introduce a design pattern that identifies
common aspects of these algorithms. It will allow us to define a strategy for
optimizing the combination of these algorithms. From the classification and a
survey of algorithms, we extract two activities common to most algorithms: the
inspection of network traffic and the evaluation of thresholds.

The former is trivial: network traffic is the carrier along which malicious
payloads reach a device directly. Indirectly a device can also detect aberrant
patterns of communication by other devices, as a result of a successful attack.

The latter is more complex: an algorithm based on signature or misuse detection
can maybe undeniably decide if an attack is real or not. But when dealing with
anomaly detection this line surely isn't that crisp.

Especially in distributed situations it is impossible to fully trust any
neighboring device. In this case one has to rely on statistics and thresholds
to decide when to drop trust in a given device.

But simply checking a threshold after updating some counter, is not a good
strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a devices
can miss a theoretical window in which one expects activity. Simply reacting on
such basic threshold can result in false positives. A way to overcome this is
to decouple the detection of problems and the decision what to do about them.

@@

We now extend the scope of the design pattern, to include its application in
the context of multiple algorithms. When implementing different algorithms, one
will typically encounter a pattern such as describes in listing
\ref{alg:id-algo-application}.

@@

\subsection*{Inherent Problems}
\label{pattern-problems}

Given these patterns, we can now identify the problem areas that tend to result
in suboptimal code when combining multiple ID algorithms. Two issues arise from
the nested loops that originate from the combination of the ID algorithm
pattern and the application pattern: loops and common data. A third issue is
the use of the network.

@@

  \item[Loops] by themselves are not a problem, but different loops, who
  perform the same operations in sequence, can incur an overhead. An example
  from the ID algorithm pattern is the handling of messages: each algorithm
  parses each observed message on the network, processing the same byte stream,
  performing comparable operations to find similar patterns.

  \item[Common data] is cross-cutting the different algorithms, that have to
  store information about nodes within its own scope. This results in a lot of
  duplicated information in memory.

@@

Reimplementing the algorithms manually to eliminate these issues is not a valid
option. This development cost typically isn't a one-time cost. For example in
case of a release of a new version of the algorithm, or simply a different
selection of algorithms due to a changing risk analysis. Also, changing, or
simply implementing the algorithms from scratch, holds the risk of making
mistakes.

The heterogeneity of IoT devices results in different software stacks, each
with for example their own network API\@. This also explains why there are no
implementations readily available to developers: there is no common platform to
target for researchers. Even using a unified software library would not only
require a consistent implementation of it, but would also limit the operational
scope of the research implementation, still restricting it to a single language.

@@

To bridge the gap between the research world and the heterogeneous IoT where
developers want to implement an IDS, while still respecting the limited
resources they have at their disposal, we introduce \NAME, a framework
consisting of a DSL, code generator and software library.

The DSL enables researchers to formally describe their ID algorithms, without
focusing on any specific language or platform. Given such formal descriptions
of ID algorithms, a code generator is now able to combine these, while avoiding
the problems arising from the patterns/problems presented in listings
\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
different languages and platforms, while reusing a software library that
provides for example message parsing, network access wrapping,\dots

Choosing for a DSL and code generator is choosing a development strategy. This
implies choices with respect to flexibility, update cost and development
effort. We will first position \NAME and motivate its design with respect to
these trade offs (\ref{positioning}). Second, we will look at the principal
design components of the DSL and how it deals with loops and common data
(\ref{dsl-design}). Third, we discuss the structure of the code generator
(\ref{code-generator-design}) and finally, we introduce the supporting software
library (\ref{software-lib-design}).

@@

Three properties classify the strategies\cite{sadilek2007energy}: flexibility,
update cost and development effort.

@@

Choosing a development strategy also determines the development cost: for example
developing for a virtual machine will be less demanding than targeting a native
image.

With \NAME we opt to start from the full native image, allowing full
flexibility, but try to reduce both the update cost and the development cost.
Firstly, by applying code generation and functional code fusion, we optimize
the code organization, which has a positive effect on the code size and hence
also on the update cost. Secondly, by enabling reuse on a functional level and
automating the writing phase, the development cost is also lowered.

\subsection{Domain Specific Language}
\label{dsl-design}

Using a DSL, and especially an external DSL, as a foundation for any solution
requires special care. It's important to remember that a DSL is merely a
\emph{thin veneer over a (semantic) model}\cite{fowler2010domain}. Its sole
purpose should be the easing of filling out the model. In the case of \NAME
this easing focuses on avoiding constructions that result in problems as
suggested in \ref{pattern-problems} concerning loops, common data and the use
of the network.

\subsubsection*{Loops}

@@

To illustrate this, we take the example of parsing network messages. Each
algorithm deals with parsing in the same way: loop over the bytes in the
payload and look for a pattern that identifies what the algorithm is looking
for. This is in fact a technical implementation choice. The underlying
functional goal is to react when such pattern is present in a message. The loop
has \emph{activated} this functionality, but the real goal is to
\emph{passively} react on an event, not \emph{actively} look for it. Event
handling can transform constructions with loops to their functional meaning,
allowing identification and fusion of common functionality.

@@

\subsubsection*{The Nodes Domain}

@@

\subsubsection*{Unified Messaging}
\label{dsl-unified-msg}

@@

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the nodes domain. An
underlying software framework will handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the nodes domain. This approach gives the
framework complete control over the messages, allowing optimized marshaling,
grouping,\dots

\subsubsection*{An Example}

Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
using a small but representative example of an algorithm: a
watchdog\cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
message at a regular interval, allowing other nodes to monitor its
availability. Missing heartbeats could indicate problems, such as physical
capturing of a device.

@@

\subsection{Code Generator}
\label{code-generator-design}

To convert the DSL-based ID algorithms into production-ready code, we propose
the use of a code generator. Although different approaches here are applicable,
like template-based text expansion, XSLT transformations or even using the
CodeDom\cite{dollard2004code}, we believe that the functional transformations
require a richer model.

A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations evolve the model into a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit source code. Figure \ref{fig:code-generation}
shows the overall procedure.

@@

\end{tikzpicture}
}
\caption{Overview code generation process}
\label{fig:code-generation}
\end{figure}

@@

\subsubsection*{Semantic Model}

@@

Figure \ref{fig:meta-model} shows the central part of the semantic model as
implemented for \NAME\@.

@@

  \draw [composite]  (nodes)    -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \NAME\@. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
\label{fig:meta-model}
\end{figure}

@@

\subsubsection*{Functional Code Fusion (FCF)}

@@

\subsubsection*{Code Model}

The code model can is comparable to an AST for an expressive language. It
contains constructs found in certain languages, but not in others. Therefore,
syntactic transformations morph the unsupported constructs into comparable
constructs supported by the targeted platform and language.

Depending on the target platform, some constructions aren't transformed into
self-containing code. The generator uses a software library for example in case
of parsing incoming messages and scheduling of tasks. It will look for patterns
that indicate any of these actions and transform them into calls into the \NAME
software library.

\subsection{Software Library}
\label{software-lib-design}

In \ref{pattern-problems} the possibility to use a software library to offload
common implementation aspects was already proposed. \ref{dsl-unified-msg}
illustrated this approach with an example use for unified messaging: parsing of
incoming messages and grouping of outgoing messages.

@@

Optimizing such algorithms is not within the scope of \NAME\@. It also gives
freedom of choice to the users of the framework to select appropriate
implementations.

\subsection{Design Summary}

As a framework \NAME tries to reduce inefficiencies in execution, memory usage
and network access. To achieve this, it tries to avoid constructions that are
hard to identify and reorganize, such as loops and common data. The DSL
enforces these restrictions.

The DSL is platform and language independent, which allows the source code
generator to target different configurations without any change to the formal
descriptions of the algorithms.

Finally, a supporting software library allows to separate functionality that is
out of scope for the generator.

@@

In this section we discuss a prototype implementation and evaluate its
performance. We first present the architecture of this implementation and
consider the impact it has from a development point of view. Next we use this
implementation to quantify its merits using a side-by-side comparison of
generated code versus a manual implementation.

\subsection*{Architecture}

@@

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform also dictates the level of effort to
implement changes to that platform. For example: to emit a different language
will require a complete rewrite of the software library and a partial rewrite
of the language description of the generator. Although this is a reusable,
one-time effort, it is important in relation to the scalability of the
framework itself.

\subsection*{Implementation}

@@

We started from a basic application that measures light intensity. The
application was an includable piece of code, which allowed integration in a
manual implementation and simple merging when generation. On top of this
baseline we added two intrusion detection related algorithms: a
watchdog\cite{mishra2004intrusion}, that allows nodes to validate each other's
presence, and a reputation-based algorithm\cite{ganeriwal2008reputation}, that
checks if a parent node is cooperative and actually forwards messages further
upstream.

\subsection*{Experimental Results}

@@

\section*{Acknowledgements}

>>>>

\input{packages}

@@

@@

\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Danny Hughes, 
Christophe Huygens, Wouter Joosen}%
\IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
3001 Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be}}

@@

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns \& anomalies that can indicate malicious
behavior. But supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-restricted devices such as those
targeting the Internet of Things. Fusion of the algorithms, to optimize sharing
of resources, can alleviate this, yet, proves to be time-consuming, repetitive
and error-prone. To address this problem we introduce \NAME, a framework for
the development of efficient intrusion detection systems. It consists of a
domain specific language, to formally describe intrusion detection algorithms
on a functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

The internet of things (IoT) holds a great potential to positively influence
our daily work and life through domotics, assisted living, e-health and
enhanced learning \cite{atzori2010internet}. But with this potential the IoT
also presents a significant threat: by opening our smart homes and personal
data to all these interconnected devices, we also open them to everyone who is
able to break the virtual locks (i.e.\ passwords, encryption keys,\dots) that
protect them. IoT devices typically have limited batteries, processing power
and memory. These properties make it hard to add any additional service,
including security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally, after detecting a malicious
pattern, alerting the firewall \cite{denning1987intrusion}.

% local ids -> localize -> limited resources

In wireless networks of resource constraint devices, which make up an
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic \cite{mishra2004intrusion}. Every
device has to implement its own lines of defense. But making the intrusion
detection system (IDS) a local service on each device requires local resources,
which are a scarce commodity for IoT devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, over the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service or access the network in a legitimate way. For each of these attacks
an algorithm needs to look for patterns or anomalies. An IDS for IoT devices
therefore requires a large number of implemented algorithms for an adequate
coverage.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. The first
aspect that is missing, is that they don't offer a way to optimize the usage of
resources. A solution that supports the integration of multiple algorithms on
IoT devices needs to consider the impact on their resources and avoid
accumulating their impact.

A second omission is support for systematic reuse to create different
configurations for different devices, both in hard- and software. Addressing
this in a transparent and automated way is important for any kind of
development in this area.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
in the implementation of these classes and details the ideas behind FCF\@.
Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation of the framework. Section
\ref{related} explores related work in the field of DSLs for ID and code
generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

@@

Two aspects, currently missing from ID frameworks, that \NAME wants to cover
are optimization of resources and support for systematic reuse. This section
uses a classification to define the scope of \NAME, detailing which types of
algorithms are good candidates for the proposed approach.

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because nodes in
the network can miss certain events that could reveal intrusions. In a wireless
network, where not all participants are within each others' radio range, nodes
typically only have a partial view on the entire network. This prohibits them
to observe potentially crucial events in distant parts of the network. The fact
that resource constraint devices typically apply duty cycling to lower their
energy consumption, further increases this problem, introducing periods of time
where they even don't see local events.

This leads to two complementary classes of ID activities: local intrusion
detection and cooperative decision making.

\subsection{Local Intrusion Detection}
\label{detection}

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors agree with this typology and add the
notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories seem promising, the authors admit that the impact on
resource-constrained devices will be too much. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Anomaly detection] considers a baseline profile and evaluates all
  activity according to this profile. It tries to flag aberrant behavior as a
  possible intrusion. The baseline profile can evolve over time, especially in
  wireless and mobile environments. Single events should therefore not result
  in absolute decisions.

  \item[Signature or misuse detection] starts with known patterns of attacks.
  It tries to find evidence for intrusions in the actions of and communication
  with a device. Because attacks often use valid ways to communicate with
  services on the devices, this approach can produce false positives.

@@

\subsection{Cooperative Decision Making}
\label{coorperative}

These taxonomies focus on the act of detecting intrusions themselves. In a
distributed environment, communication between participants may be even more
important.

Because not all devices are always actively participating in the network, they
can miss clues that would otherwise lead them to detect intrusions. It is
hardly impossible for a single device to detect a complex attack by itself.
Therefore a second important research topic consists of cooperative algorithms
to combine information from single identities into a group-based decision about
alleged intrusions. We identify two ways to construct cooperative decision
making and focus on the way participants exchange information: broadcasting and
interactive.

@@

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol \cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, the
  participants can make a distributed decision about the jointly identified
  intruder.

@@

\subsection{Properties and Scope of \NAME}
\label{scope}

Algorithms that fit this classification are typically platform independent.
They analyze network messages, looking for patterns that can reveal intrusions
and deal with the same information: other nodes in the network. To be able to
formally describe ID algorithms and target the heterogeneous IoT, \NAME
requires ID algorithms to expose these properties.

A class of ID algorithms that does not have these properties is software
attestation \cite{seshadri2008sake}. It enables exchanging information about
the software that is running on a device and identifying devices with modified
content. These algorithms are highly platform dependent and have no data in
common with other algorithms, as they look for the effects of an intrusion, not
the intruder itself.

Figure \ref{fig:classification} visually summarizes the dual classification and
scope of \NAME.

@@

Given this classification, we now introduce a design pattern that identifies
common functional aspects of these algorithms. It will allow us to define a
strategy for optimizing the combination of these algorithms. From the
classification and a survey of algorithms, we extract two activities common to
most algorithms: the inspection of network traffic and the evaluation of
thresholds.

Network traffic is the carrier along which malicious payloads reach a device
directly. Indirectly a device can also detect aberrant patterns of
communication by other devices, as a result of a successful attack.

Not all algorithms can decide if an attack is real or not, based on a single
event. Signature and misuse detection can in some cases, but anomaly detection
typically deals with some level of uncertainty. Also, in distributed situations
it is impossible to fully trust any neighboring device. In these situations one
has to rely on statistics and thresholds to decide when to drop trust in a
given device.

But simply checking a threshold after updating some counter, is not a good
strategy. A simple example is that of a watchdog \cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a device
can miss a theoretical window in which one expects activity, i.e.\ the wireless
network being the most predominantly unreliable component. Directly reacting on
a basic threshold can result in false positives. A way to overcome this is to
decouple the detection of problems and the decision what to do about them.

@@

We now extend the scope of the design pattern, to include its application in
the context of multiple algorithms. When implementing different algorithms, a
pattern such as described in listing \ref{alg:id-algo-application} will
typically appear.

@@

\subsection{Inherent Problems}
\label{pattern-problems}

These patterns can now help in identifing the problem areas that tend to result
in suboptimal code when combining multiple ID algorithms. Two issues arise from
the nested loops that originate from the combination of the ID algorithm
pattern and the application pattern: loops and common data. A third issue is
the use of the network.

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can incur an overhead. An example
  from the ID algorithm pattern is the handling of messages: each algorithm
  parses each observed message on the network, processing the same byte stream,
  performing comparable operations to find similar patterns.

  \item[Common data] cross-cuts the different algorithms, that have to store
  information about nodes within its own scope. This results in a lot of
  duplicated information in memory.

@@

\bigskip
Two operational aspects are orthogonal to these problems and amplify them:

  \item[Reimplementing] the algorithms manually to eliminate these issues is
  not a one-time cost. For example due to the release of a new version of an
  algorithm, or simply a reselection of algorithms due to a changing risk
  analysis. Also, changing, or simply implementing the algorithms from scratch,
  holds the risk of making mistakes.

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This also explains why there
  are no implementations readily available to developers: there is no single
  platform to target for researchers.

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms.
But at the same time they conflict with their target environment, the IoT\@.
\NAME provides a framework to efficiently implement an IDS for the IoT and
offers a solution for the identified problems. It consists of a DSL, source
code generator and software library. Figure \ref{fig:design} visualises the
overall design.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                 {DSL};
  \node (cg)   [representation,right=of dsl,align=center] {Code\\Generator};
  \node (code) [resource,right=of cg]                     {Code};

  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {formal};
  \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};

\end{tikzpicture}
}
\caption{Overall Design}
\label{fig:design}
\end{figure}

The DSL enables researchers to formally describe ID algorithms, without
focusing on any specific language or platform. A code generator can combine
such formal descriptions, avoiding the identified problems concerning loops,
common data and network access. It can also target different languages and
platforms, while reusing a software library that provides services including
parsing, network access wrapping,\dots

A DSL and code generator defines a development strategy. This implies choices
with respect to flexibility, update cost and development effort. We will first
position \NAME and motivate its design with respect to these trade offs.
Second, we will look at the principal design components of the DSL and how it
deals with loops and common data. Third, we discuss the structure of the code
generator and introduce its software library, that offers a way to deal with
the network access issue.

@@

Three properties can classify development strategies \cite{sadilek2007energy}:
flexibility, update cost and development effort.

@@

\NAME starts from the full native image, allowing full flexibility, and aims to
reduce both the update and development cost. Firstly, by applying functional
code fusion, \NAME optimizes the code organization, which has a positive effect
on the code size and hence also on the update cost. Secondly, by enabling reuse
on a functional level and automating the writing phase, the development cost is
also lowered.

\subsection{\NAME as a Domain Specific Language}
\label{dsl-design}

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \NAME's DSL and model focuses on avoiding constructions
that result in problems concerning loops, common data and the use of the
network. To do this, \NAME uses an event-based approach to defining
functionality.

\subsubsection{Loops}

@@

The example of parsing network messages illustrates this. Each algorithm deals
with parsing in the same way: loop over the bytes in the payload and look for a
pattern that identifies what the algorithm is looking for. This is in fact a
technical implementation choice. The underlying functional goal is to react
when such a pattern is present in a message. The loop has \emph{activated} this
functionality, but the real goal is to \emph{passively} react on an event, not
\emph{actively} look for it. Event handling therefore can transform
constructions with loops to their functional meaning, allowing identification
and fusion of common functionality.

@@

\subsubsection{The Nodes Domain}

@@

\subsubsection{Unified Messaging}
\label{dsl-unified-msg}

@@

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the nodes domain. An
underlying software framework will handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the nodes domain. This approach gives the
framework complete control over communication, allowing optimized marshaling,
grouping,\dots

\subsubsection{An Example}

Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
using a small but representative example of an algorithm: a
watchdog \cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
message at a regular interval, allowing other nodes to monitor its
availability. Missing heartbeats could indicate problems, such as physical
capturing of a device.

@@

\subsection{\NAME as a Code Generator}
\label{code-generator-design}

To convert the DSL-based ID algorithms into production-ready code, we apply
code generation techniques. Although different approaches here are applicable,
like template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the functional transformations require
a richer model.

A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations evolve the model into a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit source code. Figure \ref{fig:code-generation}
shows the process that \NAME implements.

@@

\end{tikzpicture}
}
\caption{Overview of \NAME's code generation process}
\label{fig:code-generation}
\end{figure}

@@

\subsubsection{Semantic Model}

@@

Figure \ref{fig:meta-model} shows the central part of the semantic model as
implemented for \NAME.

@@

  \draw [composite]  (nodes)    -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
\label{fig:meta-model}
\end{figure}

@@

\subsubsection{Functional Code Fusion (FCF)}

@@

\subsubsection{Code Model}

The code model is an AST for a virtual, expressive language. It contains
constructs and ideas found in different languages and programming paradigms. No
single existing programming language supports all features as is. Syntactic
transformations therefore morph the unsupported constructs into comparable
constructs supported by the targeted platform and language.

Depending on the target platform, some constructions aren't transformed into
actually corresponding code. In selected cases, the generator uses a software
library. For example when it needs to parse incoming messages or schedule
tasks. During the generation process, it looks for patterns that indicate any
of these actions and transforms them into calls to the \NAME software library.

\subsection{\NAME as a Software Library}
\label{software-lib-design}

Section \ref{pattern-problems} introduced the option to use a software library
to offload common implementation aspects, thus avoiding small discrepancies in
code and suboptimal use of shared resources. Section \ref{dsl-unified-msg}
illustrated this approach with a case for unified messaging: parsing of
incoming messages and grouping of outgoing messages.

@@

Optimizing such algorithms is not within the scope of \NAME. It also gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

\subsection{\NAME as a Framework}

\NAME reduces inefficiencies in execution, memory usage and network access, due
to the selection of separate ID algorithms. To achieve this, it avoids
constructions that are hard to identify and reorganize, such as loops and
common data. A DSL, focussing on event-handling, enforces these restrictions
and enables formally describing ID algorithms.

The DSL is also platform and language independent. This allows the source code
generator to target different configurations without any change to the formal
descriptions of the algorithms.

A supporting software library allows separation of standard functionality that
is out of scope for the generator. Optimization of the implementation of this
library for the specific platform and target language is possible.

@@

This section discusses a prototype implementation and evaluates its
performance. We first present the architecture of this implementation and
consider the impact it has from a development point of view. Next we use this
implementation to quantify its merits using a side-by-side comparison of
generated code versus a manual implementation.

\subsection{Architecture}

@@

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform also dictates the level of effort to
implement changes to that platform. For example: to generate a different
language will require a complete rewrite of the software library and a partial
rewrite of the language description of the generator. Although this is a
reusable, one-time effort, it is important in relation to the scalability of
the framework itself.

\subsection{Implementation}

@@

We started from a basic application that measures light intensity. The
application was an includable piece of code, which allowed integration in a
manual implementation and simple merging when generation. On top of this
baseline we added two intrusion detection related algorithms: a
watchdog \cite{mishra2004intrusion}, that allows nodes to validate each other's
presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
checks if a parent node is cooperative and actually forwards messages further
upstream.

\subsection{Experimental Results}

@@

\section{Acknowledgements}
